---
title: "Silicosis Diagnostic Rule"
subtitle: "Diagnostic Accuracy of Chest X-Ray for Silicosis in (Construction) Workers"
author: 
  - "Javier Mancilla Galindo, junior researcher" 
  - "Dr. Lützen Portengen, supervisor      " 
  - "Dr. Susan Peters, supervisor          "
date: today
abstract: "Objective: To perform a rapid systematic review of studies comparing chest X-Ray (index test) against chest CT (reference test) for the diagnosis of silicosis in (construction) workers."
keywords: [silicosis, ]
execute: 
  echo: false
  warning: false
format:
  html:
    toc: true
    toc_float: true
    embed-resources: true
    code-links:
        - text: "GitHub"
          href: https://github.com/UtrechtUniversity/lexces-silicosis-predict
          icon: github
  pdf: 
    toc: true
    documentclass: scrartcl
bibliography: ../docs/manuscript/lexces-silicosis-predict.bib
csl: ../docs/manuscript/american-medical-association.csl
editor: source
---

\pagebreak

```{r}
#| include: false  

# Setup 

# Create directories for sub-folders  
inputfolder <- "../data/raw"
psfolder <- "../data/processed"
tempfolder <- "../data/temp"
figfolder <- "../results/output_figures"
tabfolder <- "../results/output_tables"

dir.create(inputfolder, showWarnings = FALSE)
dir.create(psfolder, showWarnings = FALSE)
dir.create(tempfolder, showWarnings = FALSE)
dir.create(figfolder, showWarnings = FALSE)
dir.create(tabfolder, showWarnings = FALSE)
```

```{r}
#| include: false  

# Packages used 

if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(
  tidyverse,  # Used for basic data handling and visualization.
  rentrez,    # Used to search PubMed.
  XML,        # Used to parse XML data.
  gt,         # Used to print html tables.  
  report      # Used to cite packages used in this session.    
)
```

# Research Question

What is the diagnostic accuracy of chest X-Ray (index test) compared to chest CT (reference test) for the diagnosis of silicosis in (construction) workers?

## PIRD Framework

P = Population: (Construction) workers with suspected silicosis

I = Index test: Chest X-Ray

R = Reference test: Chest computed tomography (CT)

D = Disease: Silicosis or pneumoconiosis

## Terms

### Population

-   Construction workers
-   Construction industry
-   Occupational exposure
-   Occupational diseases

### Index Test

-   Chest radiography
-   Chest X-ray
-   CXR

### Reference Test

-   Computed tomography
-   CT scan
-   Chest CT

Available search block: [https://blocks.bmi-online.nl/catalog/75](https://blocks.bmi-online.nl/catalog/75)

```
"Tomography, X-Ray Computed"[Mesh] OR computed tomograph*[tiab] OR ct[tiab] OR cts[tiab] OR cat scan*[tiab] OR computer assisted tomograph*[tiab] OR computerized tomograph*[tiab] OR computerised tomograph*[tiab] OR computed x ray tomograph*[tiab] OR computed xray tomograph*[tiab]
```


### Disease

-   Silicosis
-   Pneumoconiosis


# Search Strategy

## PubMed

```{r}
#| echo: true  
query <- '(("Construction Industry"[MeSH] OR "construction"[Title/Abstract] OR "construction workers"[Title/Abstract] OR "occupational exposure"[MeSH] OR "occupational diseases"[MeSH]) AND ("Radiography, Thoracic"[MeSH] OR "chest radiography"[Title/Abstract] OR "chest X-ray"[Title/Abstract] OR "CXR"[Title/Abstract]) AND ("Tomography, X-Ray Computed"[MeSH] OR "CT scan"[Title/Abstract] OR "computed tomography"[Title/Abstract]) AND ("Silicosis"[MeSH] OR "silicosis"[Title/Abstract] OR "Pneumoconiosis"[MeSH] OR "pneumoconiosis"[Title/Abstract]))'
```


```{r}
# Search PubMed and get the count
search_results <- entrez_search(db="pubmed", term=query, retmax=0)
total_count <- search_results$count
cat("Total articles to retrieve:", total_count, "\n")
```


```{r}
#| include: false

# Initialize results dataframe
results <- data.frame(
  title = character(),
  abstract = character(),
  doi = character(),
  stringsAsFactors = FALSE
)

# Process in batches
batch_size <- 50
for(i in seq(1, total_count, batch_size)) {
  cat(sprintf("Retrieving articles %d to %d of %d\n", i, min(i + batch_size - 1, total_count), total_count))
  
  # Get batch of records
  batch_results <- entrez_search(db="pubmed", term=query, 
                                retstart=i-1, retmax=batch_size)
  
  if(length(batch_results$ids) > 0) {
    # Fetch the full records for this batch
    records <- entrez_fetch(db="pubmed", id=batch_results$ids, 
                          rettype="xml", parsed=TRUE)
    
    # Process each article in the batch
    articles <- XML::xpathApply(records, "//PubmedArticle")
    
    for(article in articles) {
      # Extract title
      title <- XML::xpathSApply(article, ".//ArticleTitle", XML::xmlValue)
      title <- if(length(title) > 0) title[1] else NA
      
      # Extract abstract
      abstract <- XML::xpathSApply(article, ".//Abstract/AbstractText", XML::xmlValue)
      abstract <- if(length(abstract) > 0) paste(abstract, collapse=" ") else NA
      
      # Extract DOI
      doi <- XML::xpathSApply(article, ".//ArticleId[@IdType='doi']", XML::xmlValue)
      doi <- if(length(doi) > 0) doi[1] else NA
      
      # Add to results
      results <- rbind(results, data.frame(
        title = title,
        abstract = abstract,
        doi = doi,
        stringsAsFactors = FALSE
      ))
    }
  }
  
  # Add a small delay to be gentle to the server
  Sys.sleep(0.5)
}
```


```{r}
# Save to CSV
write.csv(results, paste0(inputfolder,"/",lubridate::today(),"_PubMed_search.csv"), row.names = FALSE)

# Print summary
cat("\nRetrieval complete.\n")
cat("Total articles retrieved:", nrow(results), "\n")
cat("Articles with abstracts:", sum(!is.na(results$abstract)), "\n")
cat("Articles with DOIs:", sum(!is.na(results$doi)), "\n")
```

{{< pagebreak >}}



# ASReview




{{< pagebreak >}}

#### Session and package dependencies

```{r}
#| echo: false  

# remove clutter
session <- sessionInfo()
session$BLAS <- NULL
session$LAPACK <- NULL
session$loadedOnly <- NULL
# write log file
writeLines(
  capture.output(print(session, locale = FALSE)),
  paste0("sessions/",lubridate::today(), "_session_Silicosis_diagnostic_rule.txt")
)                 

session
```

{{< pagebreak >}}

# References

::: {#refs}
:::

{{< pagebreak >}}

### Package References

```{r}
#| output: asis
report::cite_packages(session)
```

```{r}
#| include: false

# Run this chunk if you wish to clear your environment.
rm(list = ls())
```
